---
title: "Final_project"
author: "Edgar Rosales,"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r}
library(caret)
```


Part 1:data importing and pre-processing



house_data, which originated from house_sales.csv and imported through the read.csv function, provides a dataset that consists of 21,613 ojbects and 21 different variables. These variables consist of numerics (int & num) and character strings.  
```{r}
house_data <- read.csv("house_sales.csv", na.strings = "")
str(house_data)
```




taking care of missing data
```{r}
colSums(is.na(house_data))
```
```{r}
house_data_clean <- na.omit(house_data)

colSums(is.na(house_data_clean))
```



checking for any duplicated rows

```{r}
#Checking for duplicated rows
duplicates <- house_data_clean[duplicated(house_data_clean), ]

#Display the duplicated rows
print(duplicates)


```



```{r}
# Convert 'date' to Date format
house_data_clean$date <- as.Date(house_data_clean$date, format="%Y%m%dT%H%M%S")
str(house_data_clean)
```




handling outliers





Here we are showing an example of what our data looks like before we handle outliers 
```{r}
# Boxplot for every numeric variable
par(mfrow = c(1, 4))  # Set the layout for multiple plots, adjust based on the number of variables
for (i in 1:ncol(house_data_clean)) {
  if (is.numeric(house_data_clean[, i])) {
    boxplot(house_data_clean[, i], main = names(house_data_clean)[i], col = "skyblue", border = "black")
  }
}
```

here we are showing the changes in boxblots after adjusting for outliers 
```{r}
#Specify the constant multiplier for outlier detection
k <- 1.5

# Loop through each numeric variable
for (col in names(house_data_clean)) {
  if (is.numeric(house_data_clean[[col]])) {
    # Calculate Q1, Q3, and IQR for the current variable
    Q1 <- quantile(house_data_clean[[col]], 0.25)
    Q3 <- quantile(house_data_clean[[col]], 0.75)
    IQR <- Q3 - Q1
    
    #Identify outliers for the current variable
    outliers <- house_data_clean[[col]] < Q1 - k * IQR | house_data_clean[[col]] > Q3 + k * IQR
    
    #Remove outliers from the dataset
    house_data_clean <- house_data_clean[!outliers, ]
  }
}

```


```{r}
par(mfrow = c(1, 4)) 
for (i in 1:ncol(house_data_clean)) {
  if (is.numeric(house_data_clean[, i])) {
    boxplot(house_data_clean[, i], main = names(house_data_clean)[i], col = "skyblue", border = "black")
  }
}
```


Here we use the aggregate function to group taverage house value per zipcode

```{r}

zip_prices <- aggregate(house_data$price, by = list(zipcode = house_data$zipcode), FUN = mean)
colnames(zip_prices) <- c("zipcode", "mean_price")

head(zip_prices)


library(ggplot2)


ggplot(zip_prices, aes(x = factor(zipcode), y = mean_price)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Average House Price by Zipcode", x = "Zipcode", y = "Average Price") +
  theme(axis.text.x = element_text(angle = 90, vjust = .5, hjust=2))



```



```{r}
house_data_clean$decade_built <- 10 * (house_data_clean$yr_built %/% 10)  # Create a decade category
avg_price_by_decade <- aggregate(house_data_clean$price, by = list(decade_built = house_data_clean$decade_built), FUN = mean)
colnames(avg_price_by_decade) <- c("decade_built", "avg_price")


barplot(avg_price_by_decade$avg_price, names.arg = avg_price_by_decade$decade_built, col = "skyblue", main = "Average House Prices by Decade of Year Built", xlab = "Decade of Year Built", ylab = "Average Price")


```



```{r}
#Aggregate average house prices by bedrooms and bathrooms
avg_price_bed_bath <- aggregate(house_data_clean$price, by = list(bedrooms = house_data_clean$bedrooms, bathrooms = house_data_clean$bathrooms), FUN = mean)
colnames(avg_price_bed_bath) <- c("bedrooms", "bathrooms", "avg_price")

#new column with formatted labels
avg_price_bed_bath$label <- paste(avg_price_bed_bath$bedrooms, "BR, ", avg_price_bed_bath$bathrooms, "BA")

#Get unique values of 'bedrooms'
unique_bedrooms <- unique(avg_price_bed_bath$bedrooms)

#separate bar charts for each unique number of bedrooms
for (bedroom_value in unique_bedrooms) {
  #Subset data for the current number of bedrooms
  subset_data <- subset(avg_price_bed_bath, bedrooms == bedroom_value)

  #new column with formatted labels
  subset_data$label <- paste(subset_data$bedrooms, "BR, ", subset_data$bathrooms, "BA")

  # Bar chart
  barplot(subset_data$avg_price, beside = TRUE, names.arg = subset_data$label, col = "lightgreen", 
          main = paste("Average House Prices by Bedrooms and Bathrooms (", bedroom_value, " Bedrooms)"), 
          xlab = "Bedrooms and Bathrooms", ylab = "Average Price (USD)")
}

```




feature construction 


```{r}
house_data_clean$age <- 2023 - house_data_clean$yr_built

```


```{r}
house_data_clean$bed_bath_ratio <- house_data_clean$bedrooms / house_data_clean$bathrooms

```


```{r}
house_data_clean$total_sqft <- house_data_clean$sqft_living + house_data_clean$sqft_lot

```


```{r}
house_data_clean$month_sold <- as.numeric(substr(house_data_clean$date, 5, 6))

```


```{r}
house_data_clean$price_per_sqft <- house_data_clean$price / house_data_clean$sqft_living

```


Normalizing
```{r}
#Function to perform Min-Max scaling
min_max_scaling <- function(x) {
  return((x - min(x)) / (max(x) - min(x)))
}

#Columns to normalize 
columns_to_normalize <- c("age", "bed_bath_ratio", "total_sqft", "month_sold", "price_per_sqft")

#Min-Max scaling to selected columns
house_data_clean[columns_to_normalize] <- lapply(house_data_clean[columns_to_normalize], min_max_scaling)

#summary statistics of normalized features
summary(house_data_clean[columns_to_normalize])

```





```{r}
# Calculate correlation matrix
correlation_matrix <- cor(house_data_clean[columns_to_normalize])

# Create a heatmap of the correlation matrix
library(ggplot2)
library(reshape2)

# Melt the correlation matrix for visualization
correlation_melted <- melt(correlation_matrix)

# Create the heatmap
ggplot(data = correlation_melted, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  theme_minimal() +
  labs(title = "Correlation Heatmap")


```






Part2

Identify categorical, ordinal and numerical veriables

```{r}
# Display the structure of the dataframe
str(house_data_clean)

```
























```{r}
# Assuming 'house_data' is your data frame
categorical_vars <- c("waterfront", "view", "zipcode")

# Checking the unique values in each categorical variable
sapply(house_data_clean[categorical_vars], function(x) unique(x))

categorical_vars

```

Identifying N

```{r}

# Assuming 'house_data' is your data frame
ordinal_vars <- c("condition", "grade")

# Checking the unique values in each ordinal variable
sapply(house_data_clean[ordinal_vars], function(x) unique(x))

```

```{r}
# Assuming 'house_data_clean' is your data frame
numerical_vars <- sapply(house_data_clean, is.numeric)
numerical_vars 

```




Measures of centraility and distribution


```{r}
# Assuming 'house_data_clean' is your data frame
numerical_vars <- sapply(house_data_clean, is.numeric)
numerical_vars <- names(numerical_vars[numerical_vars])

# Display summary statistics for numerical variables
summary_stats <- summary(house_data_clean[numerical_vars])

# Print summary statistics
print(summary_stats)

# Create histograms for numerical variables
par(mfrow = c(2, 2))  # Set up a 2x2 grid for subplots

for (var in numerical_vars) {
  hist(house_data_clean[[var]], main = paste("Histogram of", var), xlab = var, col = "lightblue", border = "black")
}


```
```{r}

correlation_matrix <- cor(house_data_clean[, sapply(house_data_clean, is.numeric)])

```



```{r}
library(caret)
nearZeroVarCols <- nearZeroVar(house_data_clean, saveMetrics = TRUE)

```




```{r}
library(caret)


nearZeroVarCols <- nearZeroVar(house_data_clean, saveMetrics = TRUE)

# Extract the names of near-zero variance columns
nearZeroVarNames <- names(house_data_clean)[nearZeroVarCols$nzv]

# Exclude near-zero variance variables from the dataset
filtered_data <- house_data_clean[, !names(house_data_clean) %in% nearZeroVarNames]



```


```{r}

# Select numeric variables for correlation analysis
numeric_variables <- filtered_data[, sapply(filtered_data, is.numeric)]

# Compute the correlation matrix for numeric variables
correlation_matrix_filtered <- cor(numeric_variables)

# Visualize the correlation matrix using a heatmap
library(corrplot)
corrplot(correlation_matrix_filtered, method = "color")

```





































